---
title: "tuneR - MIDI Demo"
author: "Taqi"
#date: "4/19/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Load dependencies
library(tuneR)
library(tidyverse)
library(wesanderson)
# Source files
# source("wrangle.R")
```

## Example

We'll be looking at the midi data for a Calvin Harris song named $``\text{blame}''$. It uses the `tuneR` package to read the midi binary into a neat dataframe of the score using the package's `getMidiNotes` function.

```{r}
# Read midi binary data
blame_midi <- readMidi("../data-midi/blame.mid")
# Get midi dataframe
blame <- getMidiNotes(blame_midi)
# Get a glimpse
head(blame)
```

## Visualizing the Score

In the end, a musical score could be (unfairly) reduced to be just a vector of notes. Harmony is just a vertical vector; melody, a horiztonal vector. We can visualize the score below.

```{r}
# Set palette
pal <- wes_palette(21, name = "GrandBudapest1", type = "continuous")
# Plot score
blame %>%
  ggplot(aes(x = time, y = note, color = notename)) +
  geom_point() +
  scale_color_manual(values = pal) +
  labs(title = "Calvin Harris - Blame (MIDI Format)")
```

